Supplement Materials of Impact of IPTp Regimen on Pregnancy Outcomes in a Malaria-Endemic Setting

### 2025-05-02 code/analysis-code/Statistical_Analysis.qmd
Project 1

library(dplyr)
library(tidyr)
library(lubridate)
library(rsample)
library(tidymodels)

# 0) Engineer features and impute their NAs
full_data <- promo_data_clean %>%
  mutate(
    enroll_month = month(enrollment_date),
    episodes_per_trimester = total_malaria_episodes /
      (gestational_age_at_enrollment_weeks / (40/3))
  )

# compute medians for imputation
median_enroll <- median(full_data$enroll_month, na.rm = TRUE)
median_eps    <- median(full_data$episodes_per_trimester, na.rm = TRUE)

full_data <- full_data %>%
  mutate(
    enroll_month           = if_else(is.na(enroll_month),    median_enroll,    enroll_month),
    episodes_per_trimester = if_else(is.na(episodes_per_trimester),
                                     median_eps,
                                     episodes_per_trimester)
  )

# 1) Safe stratification check
stratum_counts <- full_data %>% count(adverse_birth_outcome)
can_stratify   <- nrow(stratum_counts) >= 2 && all(stratum_counts$n >= 2)

# 2) 70/30 split (stratified if safe)
set.seed(1234)
split_obj <- if (can_stratify) {
  initial_split(full_data, prop = 0.7, strata = adverse_birth_outcome)
} else {
  warning("Not enough data to stratify; performing unstratified split.")
  initial_split(full_data, prop = 0.7)
}

raw_train <- training(split_obj)
raw_test  <- testing(split_obj)

# 3) Drop rows with NAs in the *core* six predictors or outcome
train_data <- raw_train %>%
  tidyr::drop_na(
    adverse_birth_outcome,
    total_malaria_episodes,
    study_arm,
    age_at_enrollment_years,
    gravidity,
    education_level
  )

test_data <- raw_test %>%
  tidyr::drop_na(
    adverse_birth_outcome,
    total_malaria_episodes,
    study_arm,
    age_at_enrollment_years,
    gravidity,
    education_level
  )

# 4) Build dynamic v-fold CV (up to 10, but ≤ nrow(train_data))
n_train <- nrow(train_data)
if (n_train < 2) stop("Not enough rows for cross‐validation (need ≥ 2).")
v <- min(10, n_train)

cv_folds <- if (can_stratify) {
  vfold_cv(train_data, v = v, strata = adverse_birth_outcome)
} else {
  vfold_cv(train_data, v = v)
}

# 5) Finally, tune your workflow—no more stray NAs in enroll_month
enet_tune <- tune_grid(
  enet_wf,
  resamples = cv_folds,
  grid      = enet_grid,
  metrics   = metric_set(roc_auc)
)

@@

# (A) Tune hyperparameters via grid search
enet_tune <- tune_grid(
  enet_wf,               # your workflow
  resamples = cv_folds,  # the vfold_cv object you just built
  grid      = enet_grid, # the grid of penalty/mixture values
  metrics   = metric_set(roc_auc)
)

# (B) Look at the tuning results
collect_metrics(enet_tune) %>% 
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean))

# (C) Pick the best parameters
best_params <- select_best(enet_tune, "roc_auc")

@@

```{r}
# 2. Read & split data
df <- read_csv(here("data","clean","PROMO_Data_clean.csv")) %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0,1)))

set.seed(2025)
split    <- initial_split(df, prop = 0.7, strata = adverse_birth_outcome)
train    <- training(split)
test     <- testing(split)
cv_folds <- vfold_cv(train, v = 5, strata = adverse_birth_outcome)

@@

```{r}
# 3. Preprocessing recipe
ml_rec <- recipe(adverse_birth_outcome ~ age_at_enrollment_years
                  + gravidity
                  + total_malaria_episodes
                  + education_level
                  + study_arm
                  + socioeconomic_status,
                data = train) %>%
  step_dummy(all_nominal_predictors()) %>%  # one‐hot encode
  step_zv(all_predictors()) %>%             # remove zero‐variance
  step_normalize(all_numeric_predictors())  # center & scale

@@

```{r}
# 6. Collect cross‐validated AUC results
lasso_auc <- lasso_res %>% filter(.metric=="roc_auc") %>% summarize(mean = mean(mean), se = mean(std_err)) %>% mutate(model="LASSO")
rf_auc    <- rf_res    %>% filter(.metric=="roc_auc") %>% summarize(mean = mean(mean), se = mean(std_err)) %>% mutate(model="RF")
xgb_auc   <- xgb_res   %>% filter(.metric=="roc_auc") %>% summarize(mean = mean(mean), se = mean(std_err)) %>% mutate(model="XGB")

bind_rows(lasso_auc, rf_auc, xgb_auc) %>%
  rename(AUC = mean, SE = se) %>%
  knitr::kable(digits=3, caption="Table 6: 5‐fold CV ROC AUC for Each Model")

@@

>>>>

library(dplyr)
library(lubridate)
library(rsample)
library(tidymodels)
library(themis)    # for step_smote()
library(scales)    # for log10_trans()

set.seed(1234)

# 1) PREPARE full_data (no complex imputation needed)
full_data <- promo_data_clean %>%
  mutate(
    adverse_birth_outcome               = factor(adverse_birth_outcome, levels = c(0,1), labels = c("no","yes")),
    total_malaria_episodes              = as.integer(total_malaria_episodes),
    gravidity                           = as.integer(gravidity),
    gestational_age_at_enrollment_weeks = as.numeric(gestational_age_at_enrollment_weeks),
    malaria_infection_rate_during_pregnancy = as.numeric(malaria_infection_rate_during_pregnancy),
    placental_malaria_by_rogerson_criteria = factor(placental_malaria_by_rogerson_criteria)
  ) %>%
  drop_na(
    adverse_birth_outcome,
    total_malaria_episodes,
    gravidity,
    gestational_age_at_enrollment_weeks,
    malaria_infection_rate_during_pregnancy,
    placental_malaria_by_rogerson_criteria,
    study_arm,
    age_at_enrollment_years,
    education_level
  )

# 2) Split 70/30 (unstratified)
split_obj  <- initial_split(full_data, prop = 0.7)
train_data <- training(split_obj)
test_data  <- testing(split_obj)

# 3) Recipe (no more enroll_month or episodes_per_trimester)
enet_recipe <- recipe(adverse_birth_outcome ~
                        total_malaria_episodes +
                        gravidity +
                        gestational_age_at_enrollment_weeks +
                        malaria_infection_rate_during_pregnancy +
                        placental_malaria_by_rogerson_criteria +
                        study_arm +
                        age_at_enrollment_years +
                        education_level,
                      data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_smote(adverse_birth_outcome) %>%
  step_normalize(all_numeric_predictors())

# 4) Model spec & workflow
enet_spec <- logistic_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

enet_wf <- workflow() %>%
  add_recipe(enet_recipe) %>%
  add_model(enet_spec)

# 5) Tuning grid
enet_grid <- grid_regular(
  penalty(range = c(-5, -1), trans = log10_trans()),
  mixture(range = c(0, 1)),
  levels = 5
)

# 6) CV folds
cv_folds <- vfold_cv(train_data, v = 5)

# 7) Tune
enet_tune <- tune_grid(
  enet_wf,
  resamples = cv_folds,
  grid      = enet_grid,
  metrics   = metric_set(roc_auc)
)

# 8) Select best & refit
best_params   <- select_best(enet_tune, metric = "roc_auc")
final_enet_wf <- finalize_workflow(enet_wf, best_params)
enet_fit      <- fit(final_enet_wf, data = train_data)

# 9) Evaluate on test set
enet_preds <- predict(enet_fit, new_data = test_data, type = "prob") %>%
  bind_cols(test_data)

print( roc_auc(enet_preds, truth = adverse_birth_outcome, .pred_yes) )

@@

# (A) Tune hyperparameters via grid search
enet_tune <- tune_grid(
  enet_wf,               # your workflow
  resamples = cv_folds,  # the vfold_cv object
  grid      = enet_grid, # grid of penalty/mixture values
  metrics   = metric_set(roc_auc)
)

# (B) Look at the tuning results for ROC AUC
roc_results <- collect_metrics(enet_tune) %>%
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean))

print(roc_results)

# (C) Pick the best parameters (note the named argument)
best_params <- select_best(enet_tune, metric = "roc_auc")

@@

```{r}
library(tidymodels)
library(readr)
library(here)

# 2. Read & split data
df <- read_csv(here("data","clean","PROMO_Data_clean.csv")) %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0,1)))

set.seed(2025)
splits    <- initial_split(df, prop = 0.7, strata = adverse_birth_outcome)
train     <- training(splits)
test      <- testing(splits)
cv_folds  <- vfold_cv(train, v = 5, strata = adverse_birth_outcome)

# 3. Preprocessing recipe
ml_rec <- recipe(adverse_birth_outcome ~ 
                   age_at_enrollment_years +
                   gravidity +
                   total_malaria_episodes +
                   education_level +
                   study_arm +
                   normalized_malaria_rate,   # ← replaced here
                 data = train) %>%
  step_dummy(all_nominal_predictors()) %>%  # one‐hot encode
  step_zv(all_predictors()) %>%             # remove zero‐variance
  step_normalize(all_numeric_predictors())  # center & scale

# 4. Model specifications & workflows

# 5. Tuning grids & 5‐fold CV
set.seed(2025)

library(dplyr)
library(knitr)

# 6) Pull the metrics into plain tibbles
lasso_metrics <- collect_metrics(lasso_res)
rf_metrics    <- collect_metrics(rf_res)
xgb_metrics   <- collect_metrics(xgb_res)

# 7) Summarize AUC for each model
lasso_auc <- lasso_metrics %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean),
    SE   = mean(std_err)
  ) %>%
  mutate(Model = "LASSO")

rf_auc <- rf_metrics %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean),
    SE   = mean(std_err)
  ) %>%
  mutate(Model = "Random Forest")

xgb_auc <- xgb_metrics %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean),
    SE   = mean(std_err)
  ) %>%
  mutate(Model = "XGBoost")

library(dplyr)
library(knitr)

# 8) Combine into one data frame and print as a plain table
table6 <- bind_rows(lasso_auc, rf_auc, xgb_auc) %>%
  dplyr::select(Model, Mean, SE)

print(table6)

@@

```{r}
library(tidymodels)  # for collect_metrics()

# 1) Collect tuning metrics into plain tibbles
lasso_metrics <- collect_metrics(lasso_res)
rf_metrics    <- collect_metrics(rf_res)
xgb_metrics   <- collect_metrics(xgb_res)

# 2) Summarize ROC AUC for each model
lasso_auc <- lasso_metrics %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean),
    SE   = mean(std_err)
  ) %>%
  mutate(Model = "LASSO")

rf_auc <- rf_metrics %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean),
    SE   = mean(std_err)
  ) %>%
  mutate(Model = "Random Forest")

xgb_auc <- xgb_metrics %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean),
    SE   = mean(std_err)
  ) %>%
  mutate(Model = "XGBoost")

# 3) Combine into one table
auc_table <- bind_rows(lasso_auc, rf_auc, xgb_auc)

# 4) Print Table 6
auc_table %>%
  dplyr::select(Model, Mean, SE) %>%
  knitr::kable(
    digits  = 3,
    caption = "Table 6: 5‑fold CV ROC AUC for Each Model"
  )
# 4) Print as a plain data frame (regular table)
print(auc_table)

```{r}
ml_rec <- recipe(adverse_birth_outcome ~
                   age_at_enrollment_years +
                   gravidity +
                   total_malaria_episodes +
                   education_level +
                   study_arm +
                   normalized_malaria_rate,
                 data = train) %>%
  step_dummy(all_nominal_predictors()) %>%   # one‑hot encode factors
  step_zv(all_predictors()) %>%              # drop zero‑variance
  step_normalize(all_numeric_predictors())   # center & scale numerics

@@

# 6) Pull the AUC metrics out and summarize
lasso_auc <- collect_metrics(lasso_res) %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean, na.rm = TRUE),
    SE   = mean(std_err, na.rm = TRUE)
  ) %>%
  mutate(Model = "LASSO")

rf_auc <- collect_metrics(rf_res) %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean, na.rm = TRUE),
    SE   = mean(std_err, na.rm = TRUE)
  ) %>%
  mutate(Model = "Random Forest")

xgb_auc <- collect_metrics(xgb_res) %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean, na.rm = TRUE),
    SE   = mean(std_err, na.rm = TRUE)
  ) %>%
  mutate(Model = "XGBoost")

# 7) Combine & print
auc_table <- bind_rows(lasso_auc, rf_auc, xgb_auc) %>%
  dplyr::select(Model, Mean, SE)

print(auc_table)

@@

*END*

### 2025-05-02 products/manuscript/Manuscript.qmd
Project 1

**Methods:**
We analyzed data from a double-blind, randomized trial in Uganda (ClinEpiDB #24, n=782 HIV-uninfected pregnant women). Participants were randomized to receive either sulfadoxine‑pyrimethamine (SP) or dihydroartemisinin‑piperaquine (DP) as IPTp. Primary outcomes were preterm birth, low birth weight, and stillbirth, combined into a composite adverse outcome. We fit multivariable logistic models evaluating (1) main effects of total malaria episodes and treatment arm, (2) their interaction, adjusting for maternal age, gravidity, and education, and (3) a prespecified subgroup analysis in women < 25 years to assess gravidity’s protective effect. Predicted probabilities were derived via ggeffects, and discriminative performance of several ML models (random forest, XGBoost, logistic regression) was summarized by AUC.

**Results:**
Adjusting for maternal age, gravidity, and socioeconomic status, our multivariable models revealed a significant interaction between malaria episode frequency and the SP regimen (OR 1.47 per additional episode, 95% CI 1.16–1.88; p = 0.002), indicating that each extra malaria episode increased the odds of an adverse birth outcome by 47% in SP‐treated women compared to those on DP. In contrast, neither malaria burden alone (OR 0.88 per episode, 95% CI 0.75–1.03; p = 0.12) nor treatment arm alone (OR 0.69 for DP vs SP, 95% CI 0.44–1.08; p = 0.10) reached significance, underscoring the importance of their interplay. In women under 25, each additional pregnancy conferred a 14% reduction in adverse‐outcome risk (OR 0.86, 95% CI 0.74–1.00; p = 0.044). The interaction model calibrated well (Hosmer–Lemeshow p = 0.37; MAE = 0.03), and on the held-out test set, the AUCs were 0.48 for logistic regression, 0.42 for the random forest, 0.44 for XGBoost, and 0.49 for the elastic net.

>>>>

**Methods:**
We analyzed data from a double‑blind, randomized trial in Uganda (ClinEpiDB#24, n=782 HIV‑uninfected pregnant women). Participants were randomized to receive either sulfadoxine-pyrimethamine (SP) or dihydroartemisinin-piperaquine (DP) as IPTp. Primary outcomes were preterm birth, low birth weight, and stillbirth, combined into a composite adverse outcome. We first fit multivariable logistic regression models to evaluate (1) the main effects of total malaria episodes and treatment arm, (2) their interaction, adjusting for maternal age, gravidity, and education level, and (3) a prespecified subgroup analysis in women <25 years to assess gravidity’s protective effect.

To explore more flexible classifiers, we compared four algorithms, standard logistic regression, elastic‑net (LASSO‑penalized) logistic regression, random forest, and XGBoost. Hyperparameters were tuned by 5‑fold cross‑validation on a 70% training split, and mean cross‑validated AUC (±SE) was used to compare methods (elastic net:0.55±0.01; random forest:0.52±0.01; XGBoost:0.50±0.01; standard logistic:0.48±0.01). The best‑tuned models were then refitted on the full training set and evaluated on the remaining 30% hold‑out set; final test‑set AUCs are reported in Table6.

**Results:**
A significant interaction between malaria episode frequency and the SP regimen emerged: each additional malaria episode under SP was associated with 47% higher odds of an adverse birth outcome than DP (OR 1.47 per episode, 95% CI 1.16–1.88; p=0.002). By itself, total malaria episodes did not significantly predict the composite outcome (OR 0.88 per episode, 95% CI 0.75–1.03; p=0.12), nor did the treatment arm alone (OR 0.69 for DP vs SP, 95% CI 0.44–1.08; p = 0.10), underscoring the critical role of their interplay. In women under 25, each additional pregnancy reduced the odds of an adverse outcome by 14% (OR 0.86, 95% CI 0.74–1.00; p=0.044). The final interaction model exhibited good calibration (Hosmer–Lemeshow p=0.37; mean absolute error=0.03). Among machine‑learning classifiers, elastic‑net (LASSO) achieved the highest mean cross‑validated AUC (0.55±0.01), followed by random forest (0.52±0.01) and XGBoost (0.50±0.01) on the 70% training split. On the independent 30% hold‑out test set, final AUCs were 0.48 for standard logistic regression, 0.42 for random forest, 0.44 for XGBoost, and 0.49 for elastic‑net (Table 6).

### 2025-05-02 products/manuscript/supplement/Supplementary-Material.qmd
Project 1

@@

>>>>

```{r}

```{r}
library(tidymodels)
library(readr)
library(here)

# 2. Read & split data
df <- read_csv(here("data","clean","PROMO_Data_clean.csv")) %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0,1)))

set.seed(2025)
splits    <- initial_split(df, prop = 0.7, strata = adverse_birth_outcome)
train     <- training(splits)
test      <- testing(splits)
cv_folds  <- vfold_cv(train, v = 5, strata = adverse_birth_outcome)

# 3. Preprocessing recipe
ml_rec <- recipe(adverse_birth_outcome ~ 
                   age_at_enrollment_years +
                   gravidity +
                   total_malaria_episodes +
                   education_level +
                   study_arm +
                   normalized_malaria_rate,   # ← replaced here
                 data = train) %>%
  step_dummy(all_nominal_predictors()) %>%  # one‐hot encode
  step_zv(all_predictors()) %>%             # remove zero‐variance
  step_normalize(all_numeric_predictors())  # center & scale

# 4. Model specifications & workflows

## (a) LASSO‐penalized logistic regression
lasso_mod <- logistic_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

lasso_wf  <- workflow() %>% 
  add_model(lasso_mod) %>% 
  add_recipe(ml_rec)

## (b) Random forest
rf_mod <- rand_forest(mtry = tune(), trees = 1000, min_n = tune()) %>%
  set_engine("ranger") %>%
  set_mode("classification")

rf_wf  <- workflow() %>% 
  add_model(rf_mod) %>% 
  add_recipe(ml_rec)

## (c) XGBoost
xgb_mod <- boost_tree(trees = tune(),
                      tree_depth = tune(),
                      learn_rate = tune()) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

xgb_wf <- workflow() %>% 
  add_model(xgb_mod) %>% 
  add_recipe(ml_rec)

# 5. Tuning grids & 5‐fold CV
set.seed(2025)

lasso_grid <- grid_regular(penalty(), levels = 20)

rf_params <- parameters(rf_mod) %>%
  update(mtry = mtry(c(1, 6)))
rf_grid   <- grid_latin_hypercube(rf_params, size = 20)

xgb_params <- parameters(xgb_mod)
xgb_grid   <- grid_latin_hypercube(xgb_params, size = 30)

ml_metrics <- metric_set(roc_auc)

lasso_res <- tune_grid(lasso_wf, resamples = cv_folds,
                       grid = lasso_grid, metrics = ml_metrics)

rf_res    <- tune_grid(rf_wf,    resamples = cv_folds,
                       grid = rf_grid,    metrics = ml_metrics)

xgb_res   <- tune_grid(xgb_wf,   resamples = cv_folds,
                       grid = xgb_grid,   metrics = ml_metrics)

library(dplyr)
library(knitr)

# 6) Pull the metrics into plain tibbles
lasso_metrics <- collect_metrics(lasso_res)
rf_metrics    <- collect_metrics(rf_res)
xgb_metrics   <- collect_metrics(xgb_res)

# 7) Summarize AUC for each model
lasso_auc <- lasso_metrics %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean),
    SE   = mean(std_err)
  ) %>%
  mutate(Model = "LASSO")

rf_auc <- rf_metrics %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean),
    SE   = mean(std_err)
  ) %>%
  mutate(Model = "Random Forest")

xgb_auc <- xgb_metrics %>%
  filter(.metric == "roc_auc") %>%
  dplyr::summarise(
    Mean = mean(mean),
    SE   = mean(std_err)
  ) %>%
  mutate(Model = "XGBoost")

library(dplyr)
library(knitr)

# 8) Combine into one data frame and print as a plain table
table6 <- bind_rows(lasso_auc, rf_auc, xgb_auc) %>%
  dplyr::select(Model, Mean, SE)

print(table6)

@@

During 5‑fold cross‑validation, the elastic‐net model achieved the highest mean AUC (0.55±0.01), followed by random forest (0.52±0.01) and XGBoost (0.50±0.01).
